---
title: "Script 02: Renaming Metabolon Lipidomics Data"
output: html_notebook
author: Raghav Jain
date: November 29, 2022
---
# 1. Summary  
The Midlife in the United States (MIDUS) study contracted Metabolon to perform lipidomic analysis. Lipidomics was performed from plasma samples collected in the MIDUS 2 (M2) and MIDUS Refresher (MR) phases. See MIDUS website for more details on the study, and the Metabolon website for information on their lipidomics platform. 

Metabolon lipid nomenclature does not follow conventions put forth by the LipidMaps Consortium & Lipidomics Standards Initiative. While these issues are unresolvable - particularly for the triglyceride species reporting - the following script will clean up column labels to be more conducive for downstream analyses. 

This script renames lipid variables from the original Metabolon dataset to make variable names more amenable to downstream processing and analysis.

Please note that the example data is randomly generated numbers. The actual lipidomics data must be obtained via request to midus.wisc.edu.
***  
# 2. Load packages and data
Hit cmd+shift+fn+f10 if on Mac, control+shift+f10 if Windows before beginning to restart R and clear previously loaded packages. 
```{r, message = F}
if(!require(tidyverse)){install.packages("tidyverse")}

# Dataset from the end of Script 01
X <- read_csv("Fake_data.csv")
```

***  
# 3. Rename variables
The lipid variable names are too long and include the broad class of lipid, then the full name of the lipid, then the abbreviation of the lipid with acyl chain information. Only the abbreviation and acyl chain are reported in working datasets.

First, the class information is deleted. That is the bolded portion of the following example:
"**Neutral Lipids: **Diacylglycerol Fatty Acid - DAG[FA18:0] microMolar"
```{r}
colnames(X) <- gsub(".*: ", "", colnames(X), ignore.case = F) # This deletes all letters up through the colon 
```

Next, the fully written out lipid species name must be deleted:
"**Diacylglycerol Fatty Acid - **DAG[FA18:0] microMolar"
```{r}
colnames(X) <- gsub(".*- ", "", colnames(X), ignore.case = F)
```

It is not convention to have 'FA' written in front of the esterified acyl chain for a given lipid - that is inferred. The follow is deleted next:
"DAG[**FA**18:0] microMolar"

Additionally, the brackets must be removed as well. 
```{r}
# A blanket command cannot be used because 'Free Fatty Acids' with the 'FFA' abbreviation are also reported. 
# Replacing 'FA' throughout the script would cause this information to be lost. Therefore, 'FFA' will be replaced by 'Free FA'
colnames(X) <- gsub("FFA", "Free FA ", colnames(X), ignore.case = F)

# Now 'FA' can be removed from all variable names and 'Free' will denote the FFAs
colnames(X) <- gsub("FA", "", colnames(X), ignore.case = F)

colnames(X) <- gsub("Free ", "FA", colnames(X), ignore.case = F)
colnames(X) <- gsub("[[]", " ", colnames(X), ignore.case = F)
colnames(X) <- gsub("]", "", colnames(X), ignore.case = F)
colnames(X) <- gsub("FA ", "FA", colnames(X), ignore.case = F)
colnames(X) <- gsub("fatty acid", "", colnames(X), ignore.case = F)
colnames(X) <- gsub("[(]", " ", colnames(X), ignore.case = F)
colnames(X) <- gsub(")", "", colnames(X), ignore.case = F)
```
Add a space after TAG with carbon numbers
```{r}
colnames(X) <- gsub("TAG", "TAG ", colnames(X), ignore.case = F)
```

For the columns with summed classes, replace the fully written out lipid names with their accepted abbreviations. Note that "Fatty acid" is already switched to "FA"; these are the free fatty acid. Nomenclature from the LipidMaps initiative will not be followed due to difficulty in distinguishing ceramide species (dihydroceramide versus ceramide). 
```{r}
colnames(X) <- gsub("Total", "Total lipid", 
                    colnames(X), 
                    ignore.case = F)   # Replace 'Total' with "Total lipid" to clarify

a <- colnames(X)[1:15]                      # Column numbers of the total lipid class variables
a <- gsub("microMolar", "", a, 
          ignore.case = F)
b <- c("CE", "MAG", "TAG", "DAG", 
       "FA", "PC", "PE", "PI", "LPC", 
       "LPE", "SM", "CER", "HCER", 
       "LCER", "DCER")

# Double check 'a' and 'b' match lipid name and abbreviation appropriately. This will be attached to final output.
Key <- data.frame(cbind(a,b))
print(Key)

for (i in 1:length(a)){
  colnames(X) <- gsub(a[i], paste0(b[i]," "), 
                      colnames(X), ignore.case = F)
}
```

Finally, remove 'microMolar' from each variable name. Output the datasheet.
```{r}
colnames(X) <- gsub(" microMolar", "", colnames(X), ignore.case = F) 
```

Output the renamed lipid data an lipid abbreviation key for MIDUS
```{r}
if(!require(openxlsx)){install.packages("openxlsx")}

wb <- createWorkbook()
addWorksheet(wb, "lipid_data")
addWorksheet(wb, "Key")

writeData(wb, "lipid_data", X, startRow = 1, startCol = 1)
writeData(wb, "Key", Key, startRow = 1, startCol = 1)

saveWorkbook(wb, file = "output/fake_data_renamed.xlsx", 
             overwrite = TRUE)
```

# I will only keep going with .csv for ease with further R processing
```{r}
write_csv(X, "output/fake_data_renamed.csv")
```
The data names have now been cleaned up. Proceed to **Script 03** for filtering of duplicate variables. 

***  
# 4. Workspace info
```{r}
sessionInfo()       # for reproducibility
```
R version 4.1.1 (2021-08-10)
Platform: x86_64-apple-darwin17.0 (64-bit)
Running under: macOS Monterey 12.3.1

Matrix products: default
LAPACK: /Library/Frameworks/R.framework/Versions/4.1/Resources/lib/libRlapack.dylib

locale:
[1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
[1] forcats_0.5.1   stringr_1.4.0   dplyr_1.0.8     purrr_0.3.4     readr_2.0.2     tidyr_1.1.4     tibble_3.1.6    ggplot2_3.3.5  
[9] tidyverse_1.3.1

loaded via a namespace (and not attached):
 [1] tidyselect_1.1.1 xfun_0.26        bslib_0.3.1      haven_2.4.3      colorspace_2.0-2 vctrs_0.3.8      generics_0.1.2   htmltools_0.5.2 
 [9] yaml_2.2.1       utf8_1.2.2       rlang_1.0.1      jquerylib_0.1.4  pillar_1.7.0     withr_2.4.3      glue_1.6.1       DBI_1.1.1       
[17] bit64_4.0.5      dbplyr_2.1.1     modelr_0.1.8     readxl_1.3.1     lifecycle_1.0.1  munsell_0.5.0    gtable_0.3.0     cellranger_1.1.0
[25] rvest_1.0.1      evaluate_0.14    knitr_1.36       tzdb_0.1.2       fastmap_1.1.0    parallel_4.1.1   fansi_1.0.2      broom_0.7.9     
[33] Rcpp_1.0.8       backports_1.2.1  scales_1.1.1     vroom_1.5.5      jsonlite_1.7.2   bit_4.0.4        fs_1.5.0         hms_1.1.1       
[41] digest_0.6.29    stringi_1.7.5    grid_4.1.1       cli_3.1.1        tools_4.1.1      magrittr_2.0.2   sass_0.4.0       crayon_1.4.2    
[49] pkgconfig_2.0.3  ellipsis_0.3.2   xml2_1.3.2       reprex_2.0.1     lubridate_1.8.0  rstudioapi_0.13  assertthat_0.2.1 rmarkdown_2.11  
[57] httr_1.4.2       R6_2.5.1         compiler_4.1.1  








