---
title: "SHOW data analysis"
author: "Raghav Jain"
date: "12/24/2022"
output:
  html_document:
    df_print: paged
---

# 1. Summary

We retrieved anthropometric data from the Survey of the Survey of the Health of Wisconsin (SHOW) study on 119 Black and White females. We performed in-house cytokine and lipidomics measurements to determine signatures of high blood pressure (HBP; systolic BP\>120 or diastolic BP\>80). Lipidomics data is in units of nM.

Release of de-identified data pertaining to this analysis and available on Github was approved by the SHOW study in accordance with their IRB. For more information on methods and the target populations, please refer to our manuscript (updated link on Github page at Rjain52/CVD-Lipidomics). \*\*\* \# 2. Load packages and data Please see 'Session info' for system requirements. For Mac, XQuartz is required to output pdf documents using 'Cairo' package. Note that all sections of the script are reliant on this section.

```{r}
# Run the packages twice to ensure installation and loading of packages
# may need to download 'xrun' package separately before loading this
if(!require(tidyverse)){install.packages("tidyverse")}
if(!require(ggpubr)){install.packages("ggpubr")}
if(!require(rstatix)){install.packages("rstatix")}
if(!require(ggsci)){install.packages("ggsci")}
if(!require(scales)){install.packages("scales")}
if(!require(Cairo)){install.packages("Cairo")}
if(!require(RColorBrewer)){install.packages("RColorBrewer")}
if(!require(ggrepel)){install.packages("ggrepel")}

# may need to download 'devEMF' and 'rvg' package separately before loading this
if(!require(webr)){install.packages("webr")}

# Set default theme for plots
theme_set(theme_bw(base_size = 16, base_family = "Arial"))

# Set default color schemes
mypal <- pal_nejm(palette = c("default"))(8)
show_col(mypal)
palette(mypal)
show_pal <- c("#20854EFF", "#E18727FF")
mypal2 <- pal_jama(palette = c("default"))(7)
palette_sc <- c("#1B1919FF", "#AD002AFF")

# Load data
my_data <- read_csv("data/SHOW_data.csv")

# Create variable to code High BP
Diag_HBP <- ifelse(my_data$`Systolic BP` > 120 | my_data$`Diastolic BP` > 80,
                             "High", "Normal")

# Add new binary BP variable as a factor
my_data <- my_data |> 
  add_column(Diag_HBP, .before = 4)

my_data$Diag_HBP <- factor(my_data$Diag_HBP, 
                           levels = c("Normal", "High"))

# Remove any participant without appropriate BP data
# This is a sanity check; no observations should be dropped for this analysis
my_data <- my_data |> drop_na(Diag_HBP) 

my_data$Race <- factor(my_data$Race, 
                           levels = c("Caucasian", 
                                      "African-American"))

my_data$Race <- recode(my_data$Race, 
                              Caucasian="White",
                             `African-American`="Black")

# Working dataset
lipid_data <- my_data
```

------------------------------------------------------------------------

# 3. General plots

Below is an example of code used for each figure.
## Cholesterol correlation plot
```{r}
# Subset Race, cholesterol and systolic BP variables
Z <- lipid_data |> select(Race, `Systolic BP`, `Total cholesterol`)

p0 <- ggscatter(Z,                        # Make scatterplot with BP on x-axis, cholesterol on y-axis
                x = "Systolic BP", 
                y = "Total cholesterol",
                add = "reg.line", conf.int = FALSE,       # Plot regression line, no confident interval
                color = "Race", palette = palette_sc,     # Add regression lines
                shape = "Race") +                         # Change point shape by Race
  stat_cor(aes(color = Race),                             # Plot Spearman correlation by Race
           method = "spearman",
           label.x = min(Z$`Systolic BP`),                # x-coordinate of label
           p.accuracy = 0.001, r.accuracy = 0.01) +       # rounding for r and p-values
  xlab("Systolic Blood Pressure (mmHg)") +                # x-axis title 
  ylab("Total cholesterol (mg/dL)") +                     # y-axis title
  ggtitle("Systolic BP versus Total Cholesterol - SHOW") +   # Plot title
  ylim(c(100, 350)) +                                     # y-axis limits
  theme_minimal() +                                       # change theme
  theme(legend.position = "right",                        # change legend, font size and font
        text = element_text(size = 16, 
                            family = "Arial"),
        axis.line = element_line(colour = "black")
        )
p0

CairoPDF(file = "output/SHOW_cholesterol_correlation.pdf", 
         height = 8, width = 10)
  print(p0)
dev.off()
```

## Violin Plot

```{r}
Z = lipid_data

# Below, the hsCRP plot is reproduced

p0 <- Z |>
  ggplot(aes(x=Race,                    # Race is the x-axis variable
             y=log10(`hsCRP (ng/mL)`),  # hsCRP is log10 transformed
             group = Diag_HBP,          # BP groups within race are plotted
             fill = Diag_HBP)) +
  geom_violin(aes(group = interaction(Diag_HBP, Race)), # violin plot added
                      scale = "area") +
  scale_fill_manual("Blood Pressure",   # Legend title and group colors
                    values = c("#709AE1FF", "#FED439FF")
                    ) +
  geom_boxplot(aes(group = interaction(Diag_HBP, Race)),  # overlay boxplot
               width = 0.3, alpha = 0.8,                  # size and color shade
               outlier.shape = NA,                        # not plotting outliers
               color = "black",                           # boxplot outline
               fill = "white",                            # boxplot fill
               position = position_dodge(width = 0.9)     # plot offset by x-axis
               ) +
  geom_point(shape = 21,                 # Add individual points
             size = 2,                
             alpha = 0.2, 
             position=position_jitterdodge(dodge.width = 0.9) # Add jitter to points
             ) + 
  theme(legend.position = "none",                 # legend plot for points
        plot.title = element_text(size = 11)
        ) +
  xlab("Race") +                                  # x-axis title
  ylab(bquote('log'[10]*'(hsCRP)')  # y-axis title
       ) +
  theme_bw(base_size = 16, base_family = "Arial") +  # simple theme
  theme(legend.position = 'right') +                 # additional theme position
  ggtitle("SHOW hsCRP (ng/mL)") +                    # plot title
  stat_compare_means(label = "p.format",             # non-parametric test
                     label.x = 1.5,
                     label.y = 1.05*(log10(max(Z$`hsCRP (ng/mL)`, 
                                               na.rm = T)
                                           )
                                     ), 
                     method = "wilcox.test")
# View plot
p0

# save .pdf of plot in already created folder titled 'output'
CairoPDF("output/SHOW_violin_CRP.pdf")
      print(p0)
dev.off()
```

## Lipidomics Methods

These are the plots describing our method.

### Scatter plot of identified lipids.

```{r}
# Read in data on identified lipids in the method
all_id <- read_csv("data/method_plot.csv")
p0 <- all_id |> ggplot(aes(x = `RT (avg)`,    # plot RT on x-axis
                           y = `Mass (avg)`,  # plot Mass on y-axis (m/z since charge always 1 for lipids)
                           color = `Class`)   # color points by lipid class
                       ) +
  geom_point() +                              # plot all lipids as points
  scale_color_manual(values = mypal2[c(1, 3:6)]) +
  ylab("mass-to-charge (mz)") +
  xlab("Retention Time (min)") +
  ggtitle("LC-MS analysis of SHOW samples", 
          subtitle = "Lipids identified in plasma."
          ) +
  theme_bw(base_size = 16) +
  theme(text = element_text(face = "bold", 
                            family = "Arial")
        )
p0

CairoPDF(file = "output/SHOW_lipidomics.pdf", 
         height = 8, width = 14)
  print(p0)
dev.off()
```

### Pie-donut chart of identified lipids.

```{r}
# Create dataframe tabulating lipids with respect to lipid class and subclass
Z <- all_id |> 
  count(Class, Abbreviation,
        sort = TRUE)            # sort from highest to lowest by number of lipids of each group

names(Z)[2] <- "Subclass"       # Rename 'Abbreviation' columns

Z |> PieDonut(aes(Class, Subclass, count=n),    # major class inside, subclass outside, size by count
                  title = "Lipid species distribution",   
                  ratioByGroup = F, 
                  explodePie = T,                  # Add white spacing between pie
                  explodeDonut = T,                # Add white spacing between donut
                  labelposition = 0)

CairoPDF(file = "output/SHOW_lipidomics_pie.pdf", 
         height = 12, width = 12)
  Z |> PieDonut(aes(Class, Subclass, count=n), 
                 title = "Lipid species distribution",
                 ratioByGroup = F,
                 explodePie = T,
                 explodeDonut = T,
                 labelposition = 0)
dev.off()
```

## Lipid class analyses

Analyze lipidomics by lipid class and acyl chain \### Class Changes in lipids Code is broken up into multiple parts. Total lipid amounts by class.

```{r}
Z <- lipid_data
Z[is.na(Z)] <- 0

# Sum total lipids by sub class based on what's in the data
Z$`Total CE` <- rowSums(Z[,grep('CE ', colnames(Z))])
Z$`Total Cer` <- rowSums(Z[,grep('Cer', colnames(Z))])
Z$`Total DG` <- rowSums(Z[,grep('\\bDG\\b', colnames(Z))])
Z$`Total EtherPC` <- rowSums(Z[,grep('EtherPC', colnames(Z))])
Z$`Total EtherPE` <- rowSums(Z[,grep('EtherPE', colnames(Z))])
Z$`Total FA` <- rowSums(Z[,grep('\\bFA\\b', colnames(Z))])
Z$`Total FFA` <- rowSums(Z[,grep('FA', colnames(Z))])  
Z$`Total LPC` <- rowSums(Z[,grep('LPC', colnames(Z))])  
Z$`Total LPE` <- rowSums(Z[,grep('LPE', colnames(Z))])  
Z$`Total PC` <- rowSums(Z[,grep('\\bPC\\b', colnames(Z))])  
Z$`Total PE` <- rowSums(Z[,grep('\\bPE\\b', colnames(Z))]) 
Z$`Total PI` <- rowSums(Z[,grep('\\bPI\\b', colnames(Z))]) 
Z$`Total SM` <- rowSums(Z[,grep('\\SM\\b', colnames(Z))]) 
Z$`Total TG` <- rowSums(Z[,grep('TG', colnames(Z))])

# Lipids used to calculate total phospholipids
my_PPL <- c("BMP", "CL", "EtherPC", "EtherPE", "LPA", "LPC", "LPE", "LPG",
            "LPI", "LPS", "PA", "PC", "PE", "PG", "PI", "PMeOH", "PS")
Z$`Total PPL` <- rowSums(Z[,grep(paste(my_PPL, collapse = "|"),
                                             colnames(Z))])

# Lipids used to calculate total (non-PPL) glycerolipids
my_GL <- c("DG","TG")
Z$`Total GL` <- rowSums(Z[,grep(paste(my_GL, collapse = "|"),
                                             colnames(Z))])

# Lipids used to calculate total sphingolipids
my_SPL <- c("Cer", "GM3", "SM")
Z$`Total SPL` <- rowSums(Z[,grep(paste(my_SPL, collapse = "|"),
                                             colnames(Z))])
# Total sterol lipids (i.e. total cholesteryl esters)
Z$`Total ST` <- Z$`Total CE`
```

```{r}
Y <- Z |> select(Simcox_ID:`TNF-a (pg/mL)`, 
                 `Total CE`:`Total ST`)

# log10 transform lipid columns
Y <- Y |> 
  mutate(across(`Total CE`:`Total ST`, log10))
```

```{r}
# Filter to target data to plot
mdat1 <- Y |> 
  filter(Race == "Black") |>                # Choose either Black or White
  select(Diag_HBP, `Total PPL`:`Total ST`)   # For major class, Total PPL:Total ST  
                                            # For subclass, Total CE:Total TG

# Make data long-form; retain HBP variable and pivot all lipid variables
mdat2 <- mdat1 |> pivot_longer(!Diag_HBP, 
                               names_to = "Lipid",
                               values_to = "value")

names(mdat2)[1] <- "Condition"             # Rename 'Diag_HBP'

# Make variables factors and change levels to desired order for plotting
mdat2$Condition <- factor(mdat2$Condition, 
                          levels = c("Normal", "High"))

mdat2$Lipid <- factor(mdat2$Lipid)
mdat2 <- na.omit(mdat2)

p0 <- ggbarplot(mdat2, x = "Lipid", y = "value",       # Create barplot
            add = c("mean_se", "dotplot"),             # elements to add
            color = "Condition",                       # outline colors; used for dotplot grouping
            fill = "Condition",                        # fill colors; used for barplot grouping
            add.params = list(width = 0.35,            # parameters for dot formatting
                              binwidth = 0.005*max(mdat2$value,     # binwidth refers to size of dots
                                                   na.rm = T)
                              ),
            alpha = 0.75,                              # opacity of colors
            position = position_dodge(0.8),            # spacing of bars w.r.t. each other 
            size = 0.15                                # width of error lines
            )                              # note: width can be used to specify individual bar size

p1 <- ggpar(p0,                        # wrapper for ggplot for aesthetics
            palette = show_pal,        # palette
            legend = "right",          # legend position 
            legend.title = "Blood Pressure") +              # legend title
  scale_y_continuous(limits = c(0.0, 1.25*max(mdat2$value)  # y-axis limits
                                ),
                     expand=c(0,0)                          # remove extra spacing on edges of plot
                     ) +       
  theme_bw(base_size = 16,                                  # various theme changes
           base_family = "Arial") +
  theme(legend.position = "right",
        text = element_text(face="bold", 
                            size = 14,
                            family = "Arial",
                            colour = "black"),
        axis.text.x = element_text(angle = 45,              # change x-axis text orientation
                                   hjust=1)                 # vertical adjustment for text
        )

# Perform t-test for pairwise comparisons
stat.test <- mdat2 |>
  group_by(Lipid) |>                  # grouping variable (x-axis)
  na.omit() |>
  t_test(value ~ Condition) |>        # perform t-test comparing lipid values by conditions (BP status)
  adjust_pvalue(method = "none") |>   # prevent automatic multiple comparison calculation
  add_significance("p")               # column for final p-value to be plotted

stat.test$p.adj.signif <- stat.test$p.signif    # Additional fail-safe to not use adjusted p-values
stat.test <- stat.test |>
  na.omit() |>
  add_xy_position(fun = "max",        # add coordinates to plot p-values based on highest point
                  "Lipid",            # x-axis variable
                  dodge = 0.8)        # spacing of p-values; should match 'position_dodge' above
  
p2 <- p1 + stat_pvalue_manual(        # add p-values to plots
  stat.test,                         
  label = "p.signif",                 # text to plot (p-value or asterisks)
  tip.length = 0.02,                  
  hide.ns = F) +                      # display non-significant (P<0.05) comparisons?
  ylab(expression(bold("log"["10"]*"(nM lipid)"))) +  # y-axis title
  xlab(expression(bold("Lipid Class"))) +             # x-axis title
  ggtitle(expression(bold("Major Class Lipids - Black")))    # plot title

p2

CairoPDF(file = "output/SHOW_Black_class_bars.pdf", 
           height = 8, width = 10)
  print(p2)
dev.off()
```

### PPL Acyl Chain comparisons

```{r}
# Filter all lipid species of a given major lipid class
Z <- lipid_data |> select(Simcox_ID, Race, Diag_HBP,      # subset these columns by name
                         grep(paste(my_PPL,               # subset all PPL lipids
                                    collapse = "|"),      # can replace with 'my_SPL' or 'my_GL'
                                  names(lipid_data)
                              )
                          )

Z <- Z |> replace(is.na(Z), 0)  # Replace NA values with 0
Z <- Z |> rename(HBP = Diag_HBP)

# Make data long-form
mdat1 <- Z |> pivot_longer(!Simcox_ID:HBP, 
                           names_to = "Lipids",
                           values_to = "values")

# Use text editing to tabulate FA intensity across lipid species
mdat1$Lipids <- gsub(".*\\ ",            # Removes characters from start '.*'
                     "", mdat1$Lipids)   # until the first space '\\ ' and replaces with nothing ('""')

# Remove the sn-2 position info on lyso-PPL
mdat1$Lipids <- gsub("/0:0", "", mdat1$Lipids)
mdat1$Lipids <- gsub("0:0/", "", mdat1$Lipids)

mdat2 <- str_split_fixed(mdat1$Lipids, "_", 2) # separate text at the '_' between FAs (change to 3 if GLs)
mdat3 <- cbind(mdat1, mdat2)

# Split individual FAs into separate columns for further editing
# Create a third data frame for GLs
mdat3_FA1 <- mdat3 |> select(Simcox_ID:values, `1`)    
mdat3_FA2 <- mdat3 |> select(Simcox_ID:values, `2`) 

names(mdat3_FA1)[6] <- "FA"
names(mdat3_FA2)[6] <- "FA"

# Replace ether lipid 'e' and duplicate annotation '_2' in FA names
mdat3_FA1$FA <- gsub("e", "", mdat3_FA1$FA)
mdat3_FA1$FA <- gsub("_2", "", mdat3_FA1$FA)
mdat3_FA2$FA <- gsub("_2", "", mdat3_FA2$FA)

# verify no other FA names need to be edited
table(mdat3_FA1$FA)
table(mdat3_FA2$FA)    # the '2' FAs refers to sum composition duplicates that can be ignored

# This is the dataset with names cleaned up
mdat4 <- rbind(mdat3_FA1, mdat3_FA2)

# Sum up acyl chain values
mdat5 <- mdat4 |> 
  group_by(Simcox_ID, Race, HBP, FA) |>  # grouping variables to retain
  summarise(value = sum(values)) |>           # function to apply (summation of value column)
  ungroup()

# Filter by Race
mdat6 <- mdat5 |> filter(Race == "Black")

# List acyl chain of interest
table(mdat6$FA)  # acyl chain options
Chains <- c("16:0", "16:1", "18:0",
            "18:1", "18:2", "20:3",
            "20:4", "22:4", "22:6")

mdat6 <- mdat6 |> filter(FA %in% Chains)  # Retain acyl chains of interest
mdat6$value <- log10(mdat6$value)         # transform data

# See previous section for details on barplot code
p0 <- ggbarplot(mdat6, x = "FA" , y = "value", 
                fill = "HBP", color = "HBP",
                add = c("mean_se", "dotplot"),  
                add.params = list(width = 0.35, 
                              binwidth = 0.005*max(mdat6$value, na.rm = T)),
                alpha = 0.75, position = position_dodge(0.8), size = 0.5
                )

p1 <- ggpar(p0, palette = show_pal, legend = "right", 
            legend.title = "Blood Pressure") +
  scale_y_continuous(limits = c(0.0, 1.25*max(mdat6$value)),
                               expand=c(0,0)
                               ) +
  theme_bw(base_size = 16, base_family = "Arial") +
  theme(legend.position = "right",
        text = element_text(face="bold",
                            size = 14,
                            family = "Arial",
                            colour = "black"),
        axis.text.x = element_text(angle = 45, hjust=1)
        )

stat.test <- mdat6 |>
  na.omit() |>
  group_by(FA) |>
  t_test(value ~ HBP) |>
  adjust_pvalue(method = "none") |>
  add_significance("p")
stat.test$p.adj.signif <- stat.test$p.signif 
stat.test <- stat.test |>
  add_xy_position(fun = "max", x = "FA", dodge = 0.8) 
  
p2 <- p1 + stat_pvalue_manual(
  stat.test, label = "p",
  tip.length = 0.02, hide.ns = F
  ) +
  ylab(expression(bold("log"["10"]*"(nM lipid)"))) +
  xlab(expression(bold("Acyl Chain"))) +
  ggtitle(expression(bold("PPL Acyl Chain Signal - Black")))

p2

CairoPDF(file = paste0("output/SHOW_PPL_Chain_Breakdown_Black.pdf"),
         height = 8, width = 10)
  print(p2)
dev.off()
```

## Lipidomics volcano plot

```{r}
# Working dataset with key grouping variables and lipids
Z <- lipid_data |>
  select(Simcox_ID, Race, Diag_HBP, 
         `CE 16:0`:`TG 58:9`)
# Make sure all 0 values are NA
Z[Z==0] <- NA

# log2 transform lipid data
Z <- Z |>
  mutate(across(c(`CE 16:0`:`TG 58:9`), log2))

# Make data long-form for target population (Black or White)
Y <- Z |> filter(Race == "White") |>             # Race filter must be applied here or FDR will be affected
  pivot_longer(cols = -c(Simcox_ID:Diag_HBP), 
                       names_to = "Lipid", 
                       values_to = "value")

# Generate data to make volcano plot
# This code performs the calculations within the group specified
volcano <- Y |> drop_na() |>
  group_by(Lipid, Race) |>      # retain Lipid names and Race info
  summarise(log2_FC = mean(value[Diag_HBP == "High"]) # Calculate fold change between high & normal BP
            - mean(value[Diag_HBP == "Normal"])) |>
  ungroup()

# Calculate p-values for high versus normal BP for each lipid, within each group
stat.test <- Y |>
  group_by(Lipid, Race) |>          # Grouping variables as above
  t_test(value ~ Diag_HBP) |>       # signifcance test (can also do wilcox_test)
  adjust_pvalue(method = "fdr") |>  # multiple comparison adjustment (adjusted values in new column)
  mutate(log10_p = -log10(p)) |>    # transform p-values for plotting
  ungroup()

# Add stats info to fold change
volcano <- left_join(volcano, stat.test, 
                     by = c("Lipid","Race"))

# Create binary [factor] variable for lipid change based on FDR p-value and direction of change
volcano$signif <- ifelse(volcano$p.adj < 0.05 & volcano$log2_FC > 0.0, 
                         "Increased", "q>0.05")
volcano$signif <- ifelse(volcano$p.adj < 0.05 & volcano$log2_FC < 0.0, 
                         "Decreased", volcano$signif)
volcano$signif <- factor(volcano$signif,
                         levels = c("q>0.05", "Increased", "Decreased"))
  
# Create scatterplot with fold change on x-axis, p-value on y-axis, and FDR adjusted direction by color
# See previous sections for explanation on ggplot options
p0 <- volcano |> ggplot(aes(x=log2_FC, y=log10_p, col=signif)) +
  geom_point() + 
  scale_x_continuous(limits = c(-0.70, max(volcano$log2_FC)*1.1)) +
  scale_y_continuous(limits = c(0, 7.50), expand = c(0.01,0)) +
  theme_minimal() +
  scale_color_manual(values=c("grey70", "#E18727FF", "#20854EFF"), 
                           name = "Change in Circulation",                   # legend title
                           breaks = c("q>0.05", "Increased", "Decreased"),   # factor levels in volcano
                           labels = c("q>0.05", "Increased", "Decreased")) + # desired name in legend
  geom_hline(yintercept = 1.301,                                             # dashed line for p<0.05
             linetype = "dashed",
             linewidth = 0.5) +
  geom_label(aes(x = 1.5, y = 1.301,           # text label for line signifying p-value
                 label = "P<0.05",
                 vjust = -0.5),
             colour = "black", label.size = NA) +
  ggtitle("SHOW - White lipid changes") +
  ylab(expression(bold("-log"["10"]*"(p-value)"))) +
  xlab(expression(bold("Fold change [log"["2"]*"(HBP) - log"["2"]*"(Normal)]"))) +
  theme_bw(base_size = 16, base_family = "Arial") + 
  theme(legend.position = "right",
        text = element_text(face="bold"),
        axis.text = element_text(colour = "black"))

p0

CairoPDF(file = "output/SHOW_White_volcano.pdf", 
         height = 8, width = 10)
 print(p0)
dev.off()
```

# 4. Multiple linear regressions

This section is broken into multiple code chunks for the different aspects of analysis.

Multiple linear regressions were performed to identify lipids significantly associated with systolic blood pressure after correcting for a number of common CVD risk factors.

Several lipid variables were randomly selected. Histograms plotting the values indicated right skewness. After testing log2 and scaling independently, it was found that log2 transformation AND scaling (pareto) gave the most normal distribution. Hence, all dependent, numerical variables were log2 transformed, then pareto scaled for linear regression.

```{r}
# Subset systolic (dependend variable), common risk factors and all lipid variables
Z <- lipid_data |>
  select(`Systolic BP`, Race, Age, `WC (cm)`, HbA1c,
         HDL, `Total cholesterol`, `hsCRP (ng/mL)`,
         `CE 16:0`:`TG 58:9`)
```

Do NOT transform or scale the dependent variable.

```{r}
# specify y-variable to be predicted
y_var <- Z$`Systolic BP`

# Race is a dummy variable where 0 = White, 1 = Black
race_var <- Z$Race

# log2 transform and scale the rest of the covariates
Age_var <- scale(log2(Z$Age), center = TRUE, scale = TRUE)
WC_var <- scale(log2(Z$`WC (cm)`), center = TRUE, scale = TRUE)
CRP_var <- scale(log2(Z$`hsCRP (ng/mL)`), center = TRUE, scale = TRUE)
Hb_var <- scale(log2(Z$HbA1c), center = TRUE, scale = TRUE)
HDL_var <- scale(log2(Z$HDL), center = TRUE, scale = TRUE)
Chol_var <- scale(log2(Z$`Total cholesterol`), center = TRUE, scale = TRUE)

# Results from the 'base' regression predicting systolic from common risk factors only
summary(lm(y_var ~ race_var + Age_var + WC_var + CRP_var + Hb_var + HDL_var + Chol_var))
# Adjusted R-squared = 0.09542, p-value = 0.01237; intercept and age are the only variables with p<0.05

# Create variables for a for loop to test the association of each variable individually
lipid <- as.character()     # variable for each lipid name

# Variables to store beta coefficient from each term 
beta_y_int <- as.numeric()
beta_AfAm <- as.numeric()
beta_Age <- as.numeric()
beta_WC <- as.numeric()
beta_CRP <- as.numeric()
beta_Hb <- as.numeric()
beta_HDL <- as.numeric()
beta_Chol <- as.numeric()
beta_lipid <- as.numeric()

# Variables to store p-value of each term in regression
p_y_int <- as.numeric()
p_AfAm <- as.numeric()
p_Age <- as.numeric()
p_WC <- as.numeric()
p_CRP <- as.numeric()
p_Hb <- as.numeric()
p_HDL <- as.numeric()
p_Chol <- as.numeric()
p_lipid <- as.numeric()

# Variable to store adjusted R-squared value of each regression
adj_r_sq <- as.numeric()

# iterate for loop starting at the first lipid variable (column 6) and ending on the last one ('ncol(Z)')
for (i in 9:ncol(Z)) {
  # For each iteration, create independent lipid variable that is log2 and scaled
  x_var <- scale(log2(Z[,i]), center = TRUE, scale = TRUE)
  
  # Run regression
  my_lm <- lm(y_var ~ race_var + Age_var + WC_var + CRP_var + Hb_var + HDL_var + Chol_var + x_var)
  
  # Store name of lipid tested 
  lipid[i] <- names(Z)[i]
  
  # store beta-coefficients; always in the first column of 'summary'
  # note: first number in bracket (row) should match position of term in regression above (intercept = 1)    
  beta_y_int[i] <- summary(my_lm)$coefficients[1,1]
  beta_AfAm[i] <- summary(my_lm)$coefficients[2,1]
  beta_Age[i] <- summary(my_lm)$coefficients[3,1]
  beta_WC[i] <- summary(my_lm)$coefficients[4,1]
  beta_CRP[i] <- summary(my_lm)$coefficients[5,1]
  beta_Hb[i] <- summary(my_lm)$coefficients[6,1]
  beta_HDL[i] <- summary(my_lm)$coefficients[7,1]
  beta_Chol[i] <- summary(my_lm)$coefficients[8,1]
  beta_lipid[i] <- summary(my_lm)$coefficients[9,1]
  
  # store term p-values; always in the fourth column of 'summary'; same note as above     
  p_y_int[i] <- summary(my_lm)$coefficients[1,4]
  p_AfAm[i] <- summary(my_lm)$coefficients[2,4]
  p_Age[i] <- summary(my_lm)$coefficients[3,4]
  p_WC[i] <- summary(my_lm)$coefficients[4,4]
  p_CRP[i] <- summary(my_lm)$coefficients[5,4]
  p_Hb[i] <- summary(my_lm)$coefficients[6,4]
  p_HDL[i] <- summary(my_lm)$coefficients[7,4]
  p_Chol[i] <- summary(my_lm)$coefficients[8,4]
  p_lipid[i] <- summary(my_lm)$coefficients[9,4]
   
  # store adjusted r-squared 
  adj_r_sq[i] <- summary(my_lm)$adj.r.sq
}

# Bind all regression variables into one matrix
final_data <- cbind(lipid, beta_y_int, beta_AfAm, beta_Age,
                    beta_WC, beta_CRP, beta_Hb, beta_HDL, beta_Chol, beta_lipid, p_y_int, 
                    p_AfAm, p_Age, p_WC, p_CRP, p_Hb, p_HDL, p_Chol, p_lipid,
                    adj_r_sq)

# Change matrix to type tibble and ensure all variables except lipid name are numeric
final_data <- as_tibble(final_data)
final_data <- final_data |> drop_na()
final_data[,2:ncol(final_data)] <- sapply(final_data[,2:ncol(final_data)],
                                          as.numeric) 
```

As stated above, the base model is significant but only Age and y-intercept are individually significant terms. This may be due collinearity between the other variables. Howver, since all current covariates are considered 'traditional' and established risk factors for CVD, we will retain them in the regressions (physiological relevance).

```{r}
# save regression info as new data frame for further manipulation
reg_plot <- final_data

# adjust lipid term p-values for multiple testing 
reg_plot$p_lipid_fdr <- p.adjust(reg_plot$p_lipid, method = 'fdr')

# Create binary term for significance (q<0.15 = Yes)
reg_plot$sig_lipid <- ifelse(reg_plot$p_lipid_fdr < 0.15, "YES", "NO")

# Create additional variable identifying if a regression adjusted r-squared > base and q<0.15
reg_plot$useful <- ifelse(reg_plot$adj_r_sq > 0.09542 & reg_plot$p_lipid_fdr < 0.15,
                          "YES", "NO")

# Create variable that will be used to plot lipid names that meet above criteria
reg_plot$delabel <- NA
reg_plot$delabel <- ifelse(reg_plot$useful == "YES",
                           reg_plot$lipid, NA)

# Retain lipid labels that correspond to target acyl chains only
toMatch <- c("20:4", "18:2", "22:4")  # specify target acyl chains (arachidonic acid related chains)

# Non-target acyl chain labels will not be labeled
reg_plot$delabel <- ifelse(grepl(paste(toMatch,collapse="|"), 
                             reg_plot$delabel),reg_plot$delabel,NA)

# Transform lipid term p-value for plotting
reg_plot$p_lipid <- (-log10(reg_plot$p_lipid))

# Create scatterplot of lipid term beta-coefficients on x-axis and significance on y-axis
# All lipids where term q<0.15 and adjusted r-squared > base regression are shaded in orange
# All arachidonic acid lipids are labeled [if meeting above criteria]
# See previous sections for explanation of ggplot functions not defined here
p0 <- reg_plot |> 
  ggplot(aes(x=beta_lipid, y=p_lipid, col=useful)) +
  geom_point(alpha = 0.5) +
  scale_x_continuous(limits = c(-6.0, 6.5)) +
  scale_y_continuous(limits = c(0, 5.0), expand = c(0.01,0)) +
  scale_color_manual(values=c("#E18727FF", "grey70"),
                     name = "Lipid term (FDR)",
                     breaks = c("YES", "NO"),
                     labels = c("q<0.15", "q>0.15")) +
  geom_text_repel(aes(label=delabel),                   # Label lpoints with arachidonic-related chain
                  color = "black",
                  segment.color = "#000000B2",
                  min.segment.length = 0, segment.size = 0.2,  # set segment size parameters
                  fontface = "bold", box.padding = 0.5,        # text parameters
                  size = 4.0, family = "Arial",                # maximum allowed label overlaps
                  max.overlaps = Inf
                  ) +
  ggtitle("SHOW Regressions - Individual Lipids") +
  ylab(expression(bold("-log"["10"]*"(p-value)"))) +
  xlab(expression(bold("Beta-coefficient"))) +
  theme_bw(base_size = 16, base_family = "Arial") +
  theme(legend.position = "right",
        text = element_text(face="bold"),
        axis.text = element_text(colour = "black")
        )
# Note, SM and HexCer are NOT ARA species; those tags can be removed later
p0

CairoPDF(file = "output/SHOW_single_lipid_regressions.pdf",
         height = 8, width = 10)
  print(p0)
dev.off()
```

# 5. 'Best' regression

Using lipids identified in the MIDUS analysis as significant predictors of systolic blood pressure, a regression predicting systolic blood pressure in the SHOW data was done.

```{r}
# Subset out key variables including systolic BP, covariates and target lipids
Z <- lipid_data |>
  select(`Systolic BP`, Race, Age, HbA1c,
         `Total cholesterol`, HDL, `hsCRP (ng/mL)`,
         `FA 18:2`, `WC (cm)`, `PC 17:0_20:3`,
         `PE 18:0_22:4`, `PE 18:0_22:4_2`,
         `EtherPE 18:1e_22:4`, `SM d36:2`,
         `PE 18:1_20:4`, `PE 16:0_18:2`,
         `TG 16:0_18:1_18:2`, `SM d38:1`, 
         `PI 18:0_18:2`)

# Log2 transform + scale all variables except systolic BP and Race 
Z[,3:ncol(Z)] <- scale(log2(Z[,3:ncol(Z)]),
                       center = T,
                       scale = T)

# Run regression
myModel <- lm(`Systolic BP`~., Z)
summary(myModel) # adjusted R2 = 0.163; p = 0.00648

# Use dependent variables to predict Systolic BP (formula set by the above model)
mypred <- predict(myModel, Z)

# Create dataframe with predicted values, actual values and Race corresponding to each observation
final <- data.frame(mypred, Z$`Systolic BP`,
                    Z$Race)

# Rename columns
colnames(final) <- c("Predicted", "Systolic", "Race")

# Create scatterplot with actual BP on x-axis and predicted BP on y-axis. Points colored by Race.
p0 <- final |>
  ggscatter(x = "Systolic", y = "Predicted",
            add = "reg.line",                 # add regression line
            color = "Race",                   # create additional grouping variable to plot one line/group
            palette = palette_sc
            ) +
  stat_cor(aes(color = Race), method = "spearman",   # Specify correlation type to report
               label.x = min(final$Systolic),     
               p.accuracy = 0.001, r.accuracy = 0.01,
               na.rm = T) +
  xlab("Actual") +
  ylab("Predicted") +
  ggtitle("SHOW Best Model - Systolic BP (mmHg)") +
  theme_bw() +
  theme(legend.position = "right",
        text = element_text(size = 16, family = "Arial"),
        axis.line = element_line(colour = "black"))

p0

# Calculate mean average error (MAE) 
final <- final[is.na(final$Predicted) == F,]
final$error = abs(final$Systolic- final$Predicted)
mean(final$error) # MAE = 13.401

CairoPDF(file = "output/SHOW_bestModel.pdf", height = 10, width = 8)
  print(p0)
dev.off()
```

# 6. System info

```{r}
sessionInfo()
```

R version 4.2.2 (2022-10-31) Platform: x86_64-apple-darwin17.0 (64-bit) Running under: macOS Monterey 12.6.2

Matrix products: default LAPACK: /Library/Frameworks/R.framework/Versions/4.2/Resources/lib/libRlapack.dylib

locale: \[1\] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8

attached base packages: \[1\] stats graphics grDevices utils datasets methods base

other attached packages: \[1\] webr_0.1.5 ggrepel_0.9.2 RColorBrewer_1.1-3 Cairo_1.6-0 scales_1.2.1\
\[6\] ggsci_2.9 rstatix_0.7.1 ggpubr_0.5.0 forcats_0.5.2 stringr_1.5.0\
\[11\] dplyr_1.0.10 purrr_0.3.5 readr_2.1.3 tidyr_1.2.1 tibble_3.1.8\
\[16\] ggplot2_3.4.0 tidyverse_1.3.2

loaded via a namespace (and not attached): \[1\] rrtable_0.3.0 googledrive_2.0.0 colorspace_2.0-3 ggsignif_0.6.4\
\[5\] rio_0.5.29 ellipsis_0.3.2 sjlabelled_1.2.0 flextable_0.8.3\
\[9\] base64enc_0.1-3 fs_1.5.2 rstudioapi_0.14 farver_2.1.1\
\[13\] bit64_4.0.5 DT_0.26 fansi_1.0.3 lubridate_1.9.0\
\[17\] xml2_1.3.3 splines_4.2.2 mnormt_2.1.1 cachem_1.0.6\
\[21\] knitr_1.41 sjmisc_2.8.9 polyclip_1.10-4 jsonlite_1.8.4\
\[25\] broom_1.0.2 dbplyr_2.2.1 ggforce_0.4.1 shiny_1.7.4\
\[29\] compiler_4.2.2 httr_1.4.4 backports_1.4.1 Matrix_1.5-3\
\[33\] assertthat_0.2.1 fastmap_1.1.0 gargle_1.2.1 cli_3.4.1\
\[37\] later_1.3.0 tweenr_2.0.2 moonBook_0.3.1 htmltools_0.5.4\
\[41\] tools_4.2.2 gtable_0.3.1 glue_1.6.2 reshape2_1.4.4\
\[45\] Rcpp_1.0.9 carData_3.0-5 jquerylib_0.1.4 cellranger_1.1.0\
\[49\] vctrs_0.5.1 nlme_3.1-161 lmtest_0.9-40 psych_2.2.9\
\[53\] insight_0.18.8 xfun_0.35.3 openxlsx_4.2.5.1 rvest_1.0.3\
\[57\] timechange_0.1.1 mime_0.12 miniUI_0.1.1.1 lifecycle_1.0.3\
\[61\] devEMF_4.1-1 googlesheets4_1.0.1 zoo_1.8-11 MASS_7.3-58.1\
\[65\] vroom_1.6.0 hms_1.1.2 promises_1.2.0.1 parallel_4.2.2\
\[69\] curl_4.3.3 sass_0.4.4 gdtools_0.2.4 stringi_1.7.8\
\[73\] zip_2.2.2 rlang_1.0.6 pkgconfig_2.0.3 systemfonts_1.0.4\
\[77\] evaluate_0.19 lattice_0.20-45 labeling_0.4.2 rvg_0.3.0\
\[81\] htmlwidgets_1.6.0 bit_4.0.5 tidyselect_1.2.0 editData_0.1.8\
\[85\] plyr_1.8.8 magrittr_2.0.3 R6_2.5.1 generics_0.1.3\
\[89\] DBI_1.1.3 mgcv_1.8-41 foreign_0.8-84 pillar_1.8.1\
\[93\] haven_2.5.1 withr_2.5.0 abind_1.4-5 ztable_0.2.3\
\[97\] modelr_0.1.10 crayon_1.5.2 car_3.1-1 shinyWidgets_0.7.5 \[101\] uuid_1.1-0 utf8_1.2.2 tzdb_0.3.0 rmarkdown_2.19\
\[105\] officer_0.5.0 grid_4.2.2 readxl_1.4.1 data.table_1.14.6\
\[109\] vcd_1.4-10 reprex_2.0.2 digest_0.6.31 xtable_1.8-4\
\[113\] httpuv_1.6.7 munsell_0.5.0 bslib_0.4.2
